---
title: "hw0317_3_23"
output:
  html_document: default
date: "2026-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
df <- read.csv("/Users/chenjunqi/Desktop/114-2/計量經濟/poe5csv/collegetown.csv")
```


```{r}
df$sqft2 <- df$sqft^2
model <- lm(price ~ sqft2, data = df)
summary(model)
```
```{r}
coef_est <- coef(summary(model))
alpha2_hat <- coef_est["sqft2", "Estimate"]
se_alpha2  <- coef_est["sqft2", "Std. Error"]
```

#a
```{r}
# 2000 square feet = 20 個「百平方英尺」
x0 <- 20

# marginal effect = dE(PRICE|SQFT)/dSQFT = 2 * alpha2 * SQFT
# 在 SQFT = 20 時，marginal effect = 40 * alpha2
marginal_hat <- 2 * alpha2_hat * x0

# H0: marginal effect <= 13
# H1: marginal effect > 13
# 因為 PRICE 單位是千美元，所以 $13000 = 13
null_value <- 13

# marginal effect 的標準誤
se_marginal <- 2 * x0 * se_alpha2
```

```{r}
# t 統計量
t_stat <- (marginal_hat - null_value) / se_marginal

# 自由度
dfree <- df.residual(model)

# 右尾檢定 p-value
p_value <- 1 - pt(t_stat, df = dfree)

# 5% 顯著水準右尾臨界值
t_crit <- qt(0.95, df = dfree)

```

```{r}
# 輸出結果
cat("alpha2_hat =", alpha2_hat, "\n")
cat("SE(alpha2_hat) =", se_alpha2, "\n")
cat("Estimated marginal effect at SQFT = 20 =", marginal_hat, "\n")
cat("SE of marginal effect =", se_marginal, "\n")
cat("t statistic =", t_stat, "\n")
cat("Critical value =", t_crit, "\n")
cat("p-value =", p_value, "\n")

# 結論
if (t_stat > t_crit) {
  cat("Reject H0: the marginal effect is greater than $13,000.\n")
} else {
  cat("Do not reject H0: insufficient evidence that the marginal effect is greater than $13,000.\n")
}
```
\[\text{Model: } PRICE = \alpha_1 + \alpha_2 SQFT^2 + e\]

The conditional expectation function is
\[E(PRICE \mid SQFT) = \alpha_1 + \alpha_2 SQFT^2.\]
Hence, the marginal effect of house size on expected price is
\[\frac{dE(PRICE \mid SQFT)}{dSQFT} = 2\alpha_2 SQFT.\]
Since \(SQFT\) is measured in hundreds of square feet, a 2000 square foot house corresponds to
\[SQFT = 20.\]
Therefore, the marginal effect at \(SQFT=20\) is
\[2\alpha_2(20) = 40\alpha_2.\]
Because \(PRICE\) is measured in thousands of dollars, \(\$13{,}000\) corresponds to \(13\).  
Thus, the hypotheses are
\[H_0: 40\alpha_2 \le 13
\qquad \text{versus}\qquad
H_1: 40\alpha_2 > 13.\]
From the regression results,
\[\hat{\alpha}_2 = 0.18452,
\qquad
se(\hat{\alpha}_2) = 0.005256.\]
Thus, the estimated marginal effect at \(SQFT=20\) is
\[40\hat{\alpha}_2 = 40(0.18452) = 7.381.\]
The standard error of \(40\hat{\alpha}_2\) is
\[se(40\hat{\alpha}_2) = 40 \cdot se(\hat{\alpha}_2)= 40(0.005256) = 0.21024.\]
So the test statistic is
\[t=\frac{40\hat{\alpha}_2 - 13}{40\,se(\hat{\alpha}_2)}=\frac{7.381 - 13}{0.21024}\approx -26.73.\]
With \(n=500\), the degrees of freedom are
\[df = 500 - 2 = 498.\]
At the 5\% significance level, this is a one-sided test, so the rejection region is
\[t > t_{0.95,498} \approx 1.648.\]
The p-value is
\[p\text{-value} = P(T_{498} > -26.73) \approx 1.000.\]
Since
\[-26.73 < 1.648,\]
we do not reject \(H_0\).
\bigskip
\textbf{Conclusion:}
There is insufficient evidence at the 5\% significance level to conclude that the marginal effect on expected house price of increasing the size of a 2000 square foot house by 100 square feet is greater than \(\$13{,}000\). The data do not support the claim.


#b

```{r}
# b 小題: 4000 square feet = 40 個「百平方英尺」
x1 <- 40

# marginal effect = dE(PRICE|SQFT)/dSQFT = 2 * alpha2 * SQFT
marginal_hat <- 2 * alpha2_hat * x1

# H0: marginal effect <= 13
# H1: marginal effect > 13
null_value <- 13

# marginal effect 的標準誤
se_marginal <- 2 * x1 * se_alpha2

# t statistic
t_stat <- (marginal_hat - null_value) / se_marginal

# 自由度
dfree <- df.residual(model)

# 右尾檢定 p-value
p_value <- 1 - pt(t_stat, df = dfree)

# 5% 顯著水準右尾臨界值
t_crit <- qt(0.95, df = dfree)

# 輸出結果
cat("Estimated marginal effect at SQFT = 40:", marginal_hat, "\n")
cat("Standard error:", se_marginal, "\n")
cat("t statistic:", t_stat, "\n")
cat("Critical value:", t_crit, "\n")
cat("p-value:", p_value, "\n")

# 結論
if (t_stat > t_crit) {
  cat("Reject H0: the marginal effect is greater than $13,000.\n")
} else {
  cat("Do not reject H0: insufficient evidence that the marginal effect is greater than $13,000.\n")
}
```
\[\text{Using the quadratic regression model }
PRICE=\alpha_1+\alpha_2 SQFT^2+e,\]
\[E(PRICE\mid SQFT)=\alpha_1+\alpha_2 SQFT^2.\]
The marginal effect is
\[\frac{dE(PRICE\mid SQFT)}{dSQFT}=2\alpha_2 SQFT.\]
Since \(SQFT\) is measured in hundreds of square feet, a 4000 square foot house corresponds to
\[SQFT=40.\]
Thus, the marginal effect at \(SQFT=40\) is
\[2\alpha_2(40)=80\alpha_2.\]
Because \(PRICE\) is measured in thousands of dollars, \(\$13{,}000\) corresponds to \(13\).  
Therefore, the hypotheses are
\[
H_0:80\alpha_2\le 13
\qquad \text{versus} \qquad
H_1:80\alpha_2>13.
\]
From part (a),
\[
\hat{\alpha}_2=0.18452,
\qquad
se(\hat{\alpha}_2)=0.005256.
\]
Hence, the estimated marginal effect at \(SQFT=40\) is
\[80\hat{\alpha}_2=80(0.18452)=14.7616.\]
Its standard error is
\[se(80\hat{\alpha}_2)=80(0.005256)=0.42048.\]
So the test statistic is
\[
t=\frac{80\hat{\alpha}_2-13}{80\,se(\hat{\alpha}_2)}
=\frac{14.7616-13}{0.42048}
\approx 4.1895.
\]

With \(n=500\), the degrees of freedom are
\[df=500-2=498.\]

At the 5\% significance level, the rejection region is
\[t>t_{0.95,498}\approx 1.648.\]

The p-value is
\[p\text{-value}=P(T_{498}>4.19)\approx 1.7\times 10^{-5}.\]

Since

\[4.1895>1.648,\]
we reject \(H_0\).

\bigskip

\textbf{Conclusion:}

At the 5\% significance level, the data support the claim that the marginal effect on expected house price of increasing the size of a 4000 square foot house by 100 square feet is greater than \(\$13{,}000\).


#c
```{r}
# 2000 square feet = 20 個「百平方英尺」
newdata <- data.frame(sqft2 = 20^2)

# 估計期望價格與 95% 信賴區間
pred_ci <- predict(model, newdata = newdata, interval = "confidence", level = 0.95)

# 印出結果
print(pred_ci)
```

\[
\text{Using the quadratic regression model }
PRICE=\alpha_1+\alpha_2 SQFT^2+e,
\]
\[E(PRICE\mid SQFT)=\alpha_1+\alpha_2 SQFT^2.\]

Since \(SQFT\) is measured in hundreds of square feet, a 2000 square foot house corresponds to

\[SQFT=20.\]

Using the estimated regression equation,

\[
\widehat{E(PRICE\mid SQFT=20)}
=
\hat{\alpha}_1+\hat{\alpha}_2(20)^2
=
93.5659+0.184519(400)
=
167.3735.
\]

Thus, the estimated expected price is

\[167.37\]

in thousand-dollar units, i.e. approximately \(\$167{,}373\).

A 95\% confidence interval for the expected price is

\[
(158.05,\;176.70)
\]

in thousand-dollar units.

Therefore, we estimate that the average price of a 2000 square foot house is about \(\$167{,}373\), and we are 95\% confident that the true expected price lies between \(\$158{,}048\) and \(\$176{,}699\).

#d
```{r}
# 找出 SQFT = 20 的房子
subset_20 <- subset(df, sqft == 20)

# 看有幾間
nrow(subset_20)

# 顯示這些房子的價格
subset_20$price

# 計算樣本平均售價
mean_price_20 <- mean(subset_20$price)
mean_price_20
```

Houses in the sample with \(SQFT=20\) have selling prices

\[138,\;169,\;183\]

(in thousand-dollar units). Therefore, the sample mean selling price is
\[\bar{PRICE}=\frac{138+169+183}{3}=163.33.\]
Thus, the sample average selling price for houses with \(SQFT=20\) is
\[\boxed{163.33}\]
in thousand-dollar units.
In part (c), the estimated expected price for a house with \(SQFT=20\) was

\[\widehat{E(PRICE\mid SQFT=20)} = 167.37\]
with a 95\% confidence interval
\[(158.05,\;176.70).\]
Since
\[163.33 \in (158.05,\;176.70),\]

the sample average selling price is compatible with the result in part (c). The observed sample mean is close to the estimated expected price and lies within the 95\% confidence interval, so there is no contradiction between the sample evidence and the model-based estimate.

