---
title: "HW0331"
author: "James"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---

# Part (e)

```{r}
set.seed(123)

N_list <- c(100, 500, 1000, 5000)

for(N in N_list){
  x <- sort(runif(N, 0, 10))
  
  mean_x <- mean(x)
  mean_x1 <- mean(x[1:(N/2)])
  mean_x2 <- mean(x[((N/2)+1):N])
  
  var_b2 <- 1000 / sum((x - mean_x)^2)
  var_b2_hat <- (4 * 1000) / (N * (mean_x2 - mean_x1)^2)
  
  sx <- sum((x - mean_x)^2) / N
  
  cat("Sample size:", N, "\n")
  cat("Mean x1, x2:", mean_x1, mean_x2, "\n")
  cat("OLS variance:", var_b2, "\n")
  cat("Mean estimator variance:", var_b2_hat, "\n")
  cat("(sx^2)^(-1):", 1/sx, "\n")
  cat("4/(x2_bar-x1_bar)^2:", 4/(mean_x2 - mean_x1)^2, "\n\n")
}
```

---

## Monte Carlo Simulation 

```{r}
simulate_one_N <- function(N, sigma2 = 1000, R = 5000) {
  half <- N/2
  
  var_b2 <- numeric(R)
  var_b2mean <- numeric(R)
  inv_sx2 <- numeric(R)
  term_mean <- numeric(R)
  
  for (r in 1:R) {
    x <- sort(runif(N, 0, 10))
    
    xbar <- mean(x)
    sx2 <- mean((x - xbar)^2)
    
    xbar1 <- mean(x[1:half])
    xbar2 <- mean(x[(half+1):N])
    
    delta <- xbar2 - xbar1
    
    var_b2[r] <- sigma2 / (N * sx2)
    var_b2mean[r] <- sigma2 * 4 / (N * delta^2)
    inv_sx2[r] <- 1 / sx2
    term_mean[r] <- 4 / (delta^2)
  }
  
  return(c(
    E_var_b2 = mean(var_b2),
    E_var_b2mean = mean(var_b2mean),
    E_inv_sx2 = mean(inv_sx2),
    E_term = mean(term_mean)
  ))
}

results <- sapply(N_list, simulate_one_N)
t(results)
```

---

## Interpretation

* Both variances decrease as (N) increases.
* OLS variance is smaller than the mean-based estimator.
* The simulation confirms:

[
E[(s_x^2)^{-1}] \approx 0.12, \quad
E\left[\frac{4}{(\bar{x}_2 - \bar{x}_1)^2}\right] \approx 0.16
]

---

# Part (f)

## Conclusion

* Both estimators have variance decreasing at rate (1/N)
* The mean-based estimator is **less efficient**
* Since:

  * it is unbiased
  * variance → 0

it appears to be **consistent**

---

# Part (h)

## Random Assignment

```{r}
simulate_h <- function(N, sigma2 = 1000, R = 5000) {
  half <- N/2
  
  term <- numeric(R)
  
  for (r in 1:R) {
    x <- runif(N, 0, 10)
    x <- sample(x)
    
    xbar1 <- mean(x[1:half])
    xbar2 <- mean(x[(half+1):N])
    
    delta <- xbar2 - xbar1
    term[r] <- 4 / (delta^2)
  }
  
  return(c(mean = mean(term), median = median(term)))
}

results_h <- sapply(c(100,500,1000), simulate_h)
t(results_h)
```

---

## Explanation

When observations are randomly assigned:

[
\bar{x}_1 \to E[x], \quad \bar{x}_2 \to E[x]
]

Thus:

[
\bar{x}_2 - \bar{x}_1 \to 0
]

This makes the estimator unstable:

[
\hat{\beta}_{2,mean}
====================

\beta_2 + \frac{\bar{e}_2 - \bar{e}_1}{\bar{x}_2 - \bar{x}_1}
]

Since the denominator goes to zero:

variance does not converge to 0
stimator is NOT consistent

---
