---
title: "HW0317"
output:
  html_document: default
  pdf_document: default
date: "2026-03-19"
---

#2.23
```{r}
library(readxl)
collegetown <- read_excel("C:/finance ecomatrics/data/collegetown.xlsx")
View(collegetown)
```


```{r}

collegetown$sqft2 <- collegetown$sqft^2
mod1 <- lm(price ~ sqft2, data = collegetown)
smod1 <- summary(mod1)
a1 <- coef(mod1)[[1]]       
a2 <- coef(mod1)[[2]]       
sea2 <- coef(smod1)[2, 2]   
df <- df.residual(mod1)   # 自由度 n-2
alpha <- 0.05
```


#(a): Marginal effect for a 2000 sq ft house

SQFT = 20 時的邊際效應 (ME) = 2 * alpha2 * 20 = 40 * alpha2
H0: 40 * alpha2 <= 13  vs  H1: 40 * alpha2 > 13
```{r}
ME_a <- 40 * a2
se_ME_a <- 40 * sea2    
#test statistic(right tail)
t_a <- (ME_a - 13) / se_ME_a
tcr_a <- qt(1 - alpha, df)  
p_value_a <- 1 - pt(t_a, df)
cat("t  =", t_a, "\ntcr =", tcr_a, "\np value =", p_value_a, "\n\n")
```
At the 5% level of significance, the calculated test statistic is $t = -26.73$, which is strictly less than the critical value $t_{cr} = 1.648$ for a right-tailed test.The $p\text{-value}$ is $1$ (greater than $0.05$)
Therefore, we do not reject the null hypothesis.
There's sufficient evidence to conclude that for a 2000-square-foot house, the marginal effect on the expected house price of adding 100 square feet is not greater than $\$13,000$

#(b): Marginal effect for a 4000 sq ft house
```{r}
ME_b <- 80 * a2
se_ME_b <- 80 * sea2        
t_b <- (ME_b - 13) / se_ME_b
tcr_b <- qt(1 - alpha, df)  
p_value_b <- 1 - pt(t_b, df)
cat("t  =", t_b, "\ntcr =", tcr_b, "\np value =", p_value_b, "\n\n")
```
At the 5% significance level, the calculated test statistic is $t = 4.189$, which strictly exceeds the critical value $t_{cr} = 1.648$, falling into the rejection region. The $p\text{-value}$ is $1.65 \times 10^{-5}$, which is highly significant and much less than $0.05$.
Therefore, we reject the null hypothesis.
There's sufficient evidence to conclude that for a 4000-square-foot house, the marginal effect of increasing its size by 100 square feet is significantly greater than $\$13,000$. 




#(c): 95% Interval estimate for expected price at SQFT=20
```{r}
new_house <- data.frame(sqft2 = 400)
ci_c <- predict(mod1, newdata = new_house, interval = "confidence", level = 0.95)
print(ci_c)
```
In repeated sampling, about 95% of those intervals to contain the true population mean. Based on our sample, the interval is $158,0481 to $176,6988.

#(d): Sample mean of prices for houses with exactly 2000 sq ft
```{r}
houses_2000 <- subset(collegetown, sqft == 20)
sample_mean_20 <- mean(houses_2000$price)
cat("實際樣本平均售價 =", sample_mean_20, "\n")
```
The actual sample mean of selling prices for a 2000-square-foot house is calculated to be $163.33$
This sample average ($163.33$) falls perfectly within the 95% confidence interval constructed in part (c), which is $[158.05, 176.70]$.


#2.31
```{r}
library(readxl)
tuna <- read_excel("C:/finance ecomatrics/data/tuna.xlsx")
View(tuna)
```


#(a): Summary Statistics & Plots
```{r}
summary(tuna$sal1)
summary(tuna$apr1)

cat("\nstandard deviation (sal1) =", sd(tuna$sal1), "\n")
cat("standard deviation (apr1) =", sd(tuna$apr1), "\n")
```


```{r}
if(!"week" %in% names(tuna)) tuna$week <- 1:nrow(tuna)
plot(tuna$week, tuna$sal1, type="l", main="Sales over time",xlab="Week", ylab="Sales")
plot(tuna$week, tuna$apr1, type="l", main="Price over time",xlab="Week", ylab="Price")
```

There is high week-to-week variation in both sales and price.

1. High Standard Deviation relative to the Mean: The standard deviation of sales (6,915) actually exceeds the sample mean (6,719). This indicates extreme dispersion in the weekly sales data.

2. From Plot: The time series plots are highly jagged and spiky, rather than smooth. The sales plot displays massive upward spikes (peaking over 30,000 units), indicating explosive, temporary growth in specific weeks.
The price plot exhibits a pattern of frequent, sharp drops followed by quick recoveries.


#(b): Scatter Plot (SAL1 vs APR1)
```{r}
plot(tuna$apr1, tuna$sal1, pch=16, main="Sales vs Price", xlab="Price", ylab="Sales")
```

The scatter plot clearly demonstrates an negative relationship between sales and price.
This result is exactly what we would expect. According to the Law of Demand, holding all else equal, as the price of a good increases, the quantity demanded decreases, vice versa.



#(c): Point & 95% Interval Estimate for 1-cent increase
```{r}
tuna$price1 <- 100 * tuna$apr1

mod2 <- lm(sal1 ~ price1, data=tuna)
smod2 <- summary(mod2)

b2 <- coef(mod2)[[2]]        
seb2 <- coef(smod2)[2,2]    

# 95% C.I. (alpha = 0.05, two-tail)
df2 <- df.residual(mod2)
tc95 <- qt(1 - 0.05/2, df2) 
lowb_c <- b2 - tc95 * seb2
upb_c <- b2 + tc95 * seb2

cat("Point Estimate (b2) =", b2, "\n")
cat("95% CI for effect of 1-cent increase = [", lowb_c, ",", upb_c, "]\n\n")
```
the estimated slope coefficient $\hat{\beta}_2$ is -434.45.
For every 1-cent increase in the price of brand no. 1 canned tuna, the expected weekly sales volume decreases by approximately 434.45 cans.

At the 95% confidence level, the interval estimate for the true marginal effect of a 1-cent price increase on sales is [-592.29, -276.60]


#(d): 90% Interval Estimate for E(sal1 | price1=70)
```{r}
tuna70 <- data.frame(price1 = 70)
predict(mod2, newdata = tuna70, interval = "confidence", level = 0.90)
```
when the price of brand no. 1 canned tuna is 70 cents/ $\$0.70$, the point estimate for the expected (average) weekly sales volume is 10,302.9 can.
 
At the 90% confidence level, the interval estimate for the true expected number of cans sold per week at a price of 70 cents is [8,624.93, 11,980.87] 


#(e) 95% Interval Estimate for Elasticity at the Means
```{r}
mean_p <- mean(tuna$price1)
mean_s <- mean(tuna$sal1)
k <- mean_p / mean_s

E_hat <- b2 * k
se_E <- seb2 * k 

tc95 <- qt(1 - 0.05/2, df2) 
lowb_e <- E_hat - tc95 * se_E
upb_e <- E_hat + tc95 * se_E

cat("Point Estimate (E_hat) =", E_hat, "\n")
cat("95% CI for effect of elasticity = [", lowb_e, ",", upb_e, "]\n\n")
```
The point estimate for the price elasticity of demand is -5.06. This indicates that a 1% increase in the price of brand no. 1 canned tuna is expected to result in approximately a 5.06% decrease in its sales volume.

At the 95% confidence level, the interval estimate for the true population elasticity evaluated at the means is [-6.90, -3.22]

Since the absolute value of the estimated elasticity is greater than 1, we conclude that the sales are fairly elastic with respect to price.

It makes perfect economic sense. A specific brand of canned tuna has many substitutes available in the same supermarket. If Brand No. 1 raises its price, consumers incur almost zero switching costs to opt for a cheaper alternative. Consequently, the quantity demanded for this specific brand is highly sensitive to any change in its price.


#(f): Hypothesis Test for Elasticity = -3
```{r}
# t statistic
t_stat <- (E_hat - (-3)) / se_E
# alpha=10% critical value
tcr_f <- qt(1 - 0.10/2, df2) 
#  p-value
p_val_f <- 2 * (1 - pt(abs(t_stat), df2))

cat("Test Statistic t =", t_stat, "\n")
cat("Critical Value tcr =", tcr_f, "\n")
cat("p-value =", p_val_f, "\n")
cat("Reject H0? ->", abs(t_stat) > tcr_f, "\n")
```
$H_0$: $E = -3$
$H_1$: $E \neq -3$

At the 10% level of significance ($\alpha = 0.10$), the critical value for a two-tailed test is $t_{cr} = 1.676$.

Rejection Region: We will reject $H_0$ if the calculated test statistic $t > 1.676$ or $t < -1.676$ ($|t| > 1.676$).

Since the test statistic $|-2.251|$ is greater than the critical value 1.676 (and the p-value 0.0288 is less than the significance level of 0.10), we fall into the rejection region. Therefore, we reject the null hypothesis.

At the 10% significance level, we have sufficient evidence to conclude that the elasticity of sales with respect to price is significantly different from -3.
