---
title: "4_25"
author: "Tony"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 4.25: Housing Price Analysis in Collegetown


```{r load_data, message=FALSE, warning=FALSE}
# Load necessary libraries
library(POE5Rdata)
library(tseries) 
library(knitr)
library(ggplot2)

# Load data
data(collegetown)
summary(collegetown)

# Calculate sample means for elasticity and slope calculations
avg_sqft <- mean(collegetown$sqft)
avg_price <- mean(collegetown$price)
```

## (a) Log-Linear Model
```{r}
mod_a <- lm(log(price) ~ sqft, data = collegetown)
summary(mod_a)

b2 <- coef(mod_a)["sqft"]

# Calculating slope and elasticity at sample means
# Slope: dP/dx = beta2 * P
# Elasticity: epsilon = beta2 * x
slope_a <- b2 * avg_price
elasticity_a <- b2 * avg_sqft
```

**Interpretation:** 
The estimated coefficient $\hat{\beta}_2$ is approximately 0.036. This suggests that for every additional 100 square feet, the house price is expected to increase by approximately 3.6%. At the sample means, the marginal effect (slope) is 9.02 and the elasticity is 0.98.


## (b) Log-Log Model
```{r}
mod_b <- lm(log(price) ~ log(sqft), data = collegetown)
summary(mod_b)

a2 <- coef(mod_b)["log(sqft)"]

# Slope and Elasticity
# Slope: dP/dx = alpha2 * (P/x)
# Elasticity: alpha2
slope_b <- a2 * (avg_price / avg_sqft)
elasticity_b <- a2
```

**Interpretation:**
The coefficient $\hat{\alpha}_2$ is 1.02, which represents a constant elasticity. A 1% increase in square footage is associated with a 1.02% increase in house price. At the sample means, the slope is 9.4.

## (c) Comparison of Goodness-of-Fit
```{r}
mod_c <- lm(price ~ sqft, data = collegetown)
summary(mod_c)

# Generalized R-squared for Log Models
y_true <- collegetown$price
gen_r2_a <- cor(y_true, exp(predict(mod_a)))^2
gen_r2_b <- cor(y_true, exp(predict(mod_b)))^2
r2_c <- summary(mod_c)$r.squared
```

**Answer:**
Comparing the generalized $R^2$, the Log-Linear model 0.6622 and Log-Log model 0.6445 both provide a better fit to the data than the simple linear model 0.6413.


## (d) Normality of Residuals
```{r}
hist(residuals(mod_a),
     breaks = 20,
     col = "#c7bf7f",
     border = "white",
     main = "Log-Linear Residuals",
     xlab = "Residuals",
     xaxs = "i",
     yaxs = "i")
box(bty = "l")

hist(residuals(mod_b),
     breaks = 20,
     col = "#c7bf7f",
     border = "white",
     main = "Log-Log Residuals",
     xlab = "Residuals",
     xaxs = "i",
     yaxs = "i")
box(bty = "l")

hist(residuals(mod_c),
     breaks = 25,
     col = "#c7bf7f",
     border = "white",
     main = "Linear Residuals",
     xlab = "Residuals",
     xaxs = "i",
     yaxs = "i")
box(bty = "l")

# Jarque-Bera Tests
jb_a <- jarque.bera.test(residuals(mod_a))
jb_b <- jarque.bera.test(residuals(mod_b))
jb_c <- jarque.bera.test(residuals(mod_c))
```

**Answer:**
The Jarque-Bera test p-values are extremely low for all models, especially for the linear model ($p \approx 0$). This indicates that the residuals are not normally distributed. However, the log-transformed models show less extreme skewness in the histograms.

## (e) Residual Patterns and Heteroscedasticity
```{r}
plot(collegetown$sqft, residuals(mod_a), main = "Log-Linear: Residual vs SQFT", xlab = "SQFT", ylab = "Residuals"
     ,pch = 19, col = "#2b4a6f", cex = 0.8)
abline(h = 0, col = "#a21313", lwd = 2)
plot(collegetown$sqft, residuals(mod_b), main = "Log-Log: Residual vs SQFT", xlab = "SQFT", ylab = "Residuals"
     ,pch = 19, col = "#2b4a6f", cex = 0.8)
abline(h = 0, col = "#a21313", lwd = 2)
plot(collegetown$sqft, residuals(mod_c), main = "Linear: Residual vs SQFT", xlab = "SQFT", ylab = "Residuals"
     ,pch = 19, col = "#2b4a6f", cex = 0.8)
abline(h = 0, col = "#a21313", lwd = 2)
```

**Answer:**
The linear model's residuals show a distinct fan shape, indicating heteroscedasticity. The log-transformed models (a) and (b) help stabilize this variance, though some patterns may still remain.

## (f)(g) Prediction for a 2700 sqft House
```{r}

new_data <- data.frame(sqft = 27) # 2700 sqft corresponds to sqft = 27 in this dataset

# Predictions (Exp-transformed for Log models)
pred_a_log <- predict(mod_a, new_data, interval = "prediction", level = 0.95)
pred_a <- exp(pred_a_log) 

pred_b_log <- predict(mod_b, new_data, interval = "prediction", level = 0.95)
pred_b <- exp(pred_b_log)

pred_c <- predict(mod_c, new_data, interval = "prediction", level = 0.95)

# Summary Table
results_comparison <- data.frame(
  Model = c("Log-Linear", "Log-Log", "Linear"),
  R_squared = c(gen_r2_a, gen_r2_b, r2_c),
  JB_p_value = c(jb_a$p.value, jb_b$p.value, jb_c$p.value),
  Pred_Price_2700 = c(pred_a[1], pred_b[1], pred_c[1])
)

kable(results_comparison, caption = "Comparison of Model Results and Predictions", digits = 4)
```

## (h) Conclusion on Functional Form
**Answer:**
Based on the analysis, the Log-Log model (b) is generally preferred. It addresses the non-linearity between house size and price more effectively than the linear model, yields a higher generalized $R^2$, and provides a more intuitive interpretation of the relationship through constant elasticity. Furthermore, it helps mitigate the heteroscedasticity observed in the original price-level residuals.

