---
title: "HW0505_314707031"
author: "James"
output: pdf_document
---

```{r setup, include=FALSE}
Sys.setenv(LANGUAGE = "en")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r packages}
library(AER)
library(car)
library(POE5Rdata)
library(knitr)
```

# Chapter 11: Simultaneous Equations

# Exercise 11.24: Fulton Fish Market

This exercise revisits the supply and demand model for fish at the Fulton Fish Market (Example 11.2), using the data file `fultonfish`. The structural demand equation (11.13) and supply equation (11.14) are

$$\ln(QUAN_t)=\alpha_1+\alpha_2\ln(PRICE_t)+\alpha_3 MON_t+\alpha_4 TUE_t+\alpha_5 WED_t+\alpha_6 THU_t+e_{dt}$$
$$\ln(QUAN_t)=\beta_1+\beta_2\ln(PRICE_t)+\beta_3 STORMY_t+e_{st}$$

In this data set `lprice` is $\ln(PRICE)$ and `lquan` is $\ln(QUAN)$. We add the indicator `mixed` (poor but not stormy weather) to the supply equation.

```{r load-fish}
data("fultonfish")
```

## 11.24(a)

```{r fish-a}
# Reduced form for ln(PRICE), equation (11.16) augmented with MIXED
rf_price <- lm(lprice ~ mon + tue + wed + thu + stormy + mixed,
               data = fultonfish)
summary(rf_price)

# Is MIXED significant at the 5% level?
coef(summary(rf_price))["mixed", ]

# Joint significance of STORMY and MIXED
linearHypothesis(rf_price, c("stormy = 0", "mixed = 0"))
```

**Explanation.**
The reduced-form equation for `lprice` now includes the two weather indicators `stormy` and `mixed` as the supply shift variables. The individual t-test shows whether `mixed` is significant at the 5% level (look at its p-value). The joint F test for `stormy = mixed = 0` is the relevance test for the instruments used to identify the demand equation. If the F-value exceeds 10, the instruments are strong enough that two-stage least squares of the demand equation can be trusted.

## 11.24(b)

```{r fish-b}
# Demand equation (11.13) estimated by 2SLS using STORMY and MIXED as IVs
demand_iv <- ivreg(lquan ~ lprice + mon + tue + wed + thu |
                     mon + tue + wed + thu + stormy + mixed,
                   data = fultonfish)
summary(demand_iv)
```

**Explanation.**
Here the demand equation is estimated by two-stage least squares, treating `lprice` as endogenous and using `stormy` and `mixed` as the excluded instruments. The coefficient on `lprice` is the price elasticity of demand, which is expected to be negative. Compared with Table 11.5 (which uses only `stormy`), adding `mixed` as a second instrument leaves the elasticity estimate close to the original value but can change its precision.

## 11.24(c)

```{r fish-c}
# Sargan test of the overidentifying restriction (validity of the surplus IV)
summary(demand_iv, diagnostics = TRUE)$diagnostics
```

**Explanation.**
Because there are two instruments (`stormy`, `mixed`) for one endogenous variable (`lprice`), the demand equation is overidentified by one restriction, so the validity of the "surplus" instrument can be tested. The Sargan test regresses the 2SLS residuals on all instruments; under the null all instruments are valid. A large p-value (failure to reject) means there is no evidence that `mixed` is an invalid instrument.

## 11.24(d)

```{r fish-d}
# In the reduced form from (a), test joint significance of the day indicators
linearHypothesis(rf_price,
                 c("mon = 0", "tue = 0", "wed = 0", "thu = 0"))
```

**Explanation.**
To estimate the **supply** equation by 2SLS we need the demand-side shift variables — the daily indicators — to be jointly significant in the reduced form for `lprice`. The F test for `mon = tue = wed = thu = 0` checks this. As is typical with this data set, the day-of-week indicators are jointly insignificant, so they are weak instruments for `lprice`. Consequently we **cannot** estimate the supply equation by 2SLS with confidence, even though the demand equation can be estimated reliably.

# Exercise 11.28: Truffle Supply and Demand with Price on the Left

This exercise uses the `truffles` data (Example 11.1). The original demand and supply equations (11.11) and (11.12) are

$$Q_i=\alpha_1+\alpha_2 P_i+\alpha_3 PS_i+\alpha_4 DI_i+e_{di}, \qquad
Q_i=\beta_1+\beta_2 P_i+\beta_3 PF_i+e_{si}$$

```{r load-truffle}
data("truffles")
```

## 11.28(a)

Rewriting the demand and supply equations with $P$ on the left-hand side gives the inverse system

$$\text{Demand: } P_i=\alpha_1^{*}+\alpha_2^{*}Q_i+\alpha_3^{*}PS_i+\alpha_4^{*}DI_i+u_{di}$$
$$\text{Supply: } P_i=\beta_1^{*}+\beta_2^{*}Q_i+\beta_3^{*}PF_i+u_{si}$$

**Anticipated signs.**
In the inverse **demand** equation the slope $\alpha_2^{*}$ should be **negative** (a downward-sloping demand curve), while $\alpha_3^{*}$ (price of the substitute $PS$) and $\alpha_4^{*}$ (income $DI$) should be **positive**, because both shift demand outward and raise price. In the inverse **supply** equation the slope $\beta_2^{*}$ should be **positive** (an upward-sloping supply curve), while $\beta_3^{*}$ (price of the factor $PF$) should be **negative**: a higher input cost shifts supply back, so at any given quantity a higher price is required, but holding $Q$ fixed the coefficient on $PF$ is expected to be negative.

## 11.28(b)

```{r truffle-b}
# Inverse demand by 2SLS: Q endogenous, IVs are the exogenous PS, DI, PF
inv_demand <- ivreg(p ~ q + ps + di | ps + di + pf, data = truffles)
summary(inv_demand)

# Inverse supply by 2SLS: Q endogenous, IVs are PS, DI, PF
inv_supply <- ivreg(p ~ q + pf | ps + di + pf, data = truffles)
summary(inv_supply)
```

**Explanation.**
Both inverse equations are estimated by two-stage least squares, treating quantity `q` as the endogenous regressor and using the exogenous variables `ps`, `di`, and `pf` as instruments. The estimated slope on `q` is negative in the demand equation and positive in the supply equation, consistent with downward-sloping demand and upward-sloping supply, and the coefficients are generally significantly different from zero.

## 11.28(c)

```{r truffle-c}
# Price elasticity of demand "at the means" from the inverse demand estimates
a2 <- coef(inv_demand)["q"]      # dP/dQ along the inverse demand curve
Pbar <- mean(truffles$p)
Qbar <- mean(truffles$q)

# elasticity of demand = (dQ/dP)*(P/Q) = (1/a2)*(Pbar/Qbar)
elas_demand <- (1 / a2) * (Pbar / Qbar)
cat("dP/dQ (slope) =", a2, "\n")
cat("P-bar =", Pbar, "  Q-bar =", Qbar, "\n")
cat("Price elasticity of demand at the means =", elas_demand, "\n")
```

**Explanation.**
The inverse demand equation gives $dP/dQ=\hat\alpha_2^{*}$, so $dQ/dP=1/\hat\alpha_2^{*}$. The price elasticity of demand at the means is $(dQ/dP)\times(\bar P/\bar Q)$. The result is about $-1.27$, i.e. negative as expected and (in absolute value) greater than one, indicating that truffle demand is price elastic at the sample means.

## 11.28(d)

```{r truffle-d, fig.width=6, fig.height=5}
# Evaluate at the given exogenous values
DI_star <- 3.5; PF_star <- 23; PS_star <- 22

bd <- coef(inv_demand)   # inverse demand coefficients
bs <- coef(inv_supply)   # inverse supply coefficients

# P as a function of Q for each curve, holding exogenous variables fixed
demand_P <- function(Q) bd["(Intercept)"] + bd["q"]*Q + bd["ps"]*PS_star + bd["di"]*DI_star
supply_P <- function(Q) bs["(Intercept)"] + bs["q"]*Q + bs["pf"]*PF_star

Qseq <- seq(0, 40, length.out = 200)
plot(Qseq, demand_P(Qseq), type = "l", col = "blue", lwd = 2,
     xlab = "Q (quantity)", ylab = "P (price)",
     main = "Truffle Supply and Demand (P on vertical axis)")
lines(Qseq, supply_P(Qseq), col = "red", lwd = 2)
legend("topright", legend = c("Demand", "Supply"),
       col = c("blue", "red"), lwd = 2)
```

**Explanation.**
The curves are drawn with $P$ on the vertical axis and $Q$ on the horizontal axis, using $DI^{*}=3.5$, $PF^{*}=23$, $PS^{*}=22$. The demand curve slopes downward and the supply curve slopes upward, intersecting at the market equilibrium.

## 11.28(e)

```{r truffle-e}
# Equilibrium from part (d): set demand_P(Q) = supply_P(Q) and solve for Q
# (bd0 + bd_q*Q) = (bs0 + bs_q*Q)  with intercepts already including exog terms
bd0 <- bd["(Intercept)"] + bd["ps"]*PS_star + bd["di"]*DI_star
bs0 <- bs["(Intercept)"] + bs["pf"]*PF_star

Q_eq <- (bs0 - bd0) / (bd["q"] - bs["q"])
P_eq <- demand_P(Q_eq)
cat("Equilibrium from structural curves:  Q =", Q_eq, " P =", P_eq, "\n")

# Compare with reduced-form predictions (Table 11.2): regress P and Q on all exogenous
rf_Q <- lm(q ~ ps + di + pf, data = truffles)
rf_P <- lm(p ~ ps + di + pf, data = truffles)
newx <- data.frame(ps = PS_star, di = DI_star, pf = PF_star)
Q_rf <- predict(rf_Q, newx)
P_rf <- predict(rf_P, newx)
cat("Equilibrium from reduced forms:      Q =", Q_rf, " P =", P_rf, "\n")
```

**Explanation.**
Setting the two estimated inverse curves equal gives the equilibrium $(Q,P)$ implied by the structural 2SLS estimates. The reduced-form equations predict $P$ and $Q$ directly from the exogenous variables at the same values $DI^{*},PF^{*},PS^{*}$. The two sets of equilibrium values are very close, confirming the internal consistency between the structural and reduced-form representations.

## 11.28(f)

```{r truffle-f}
# OLS of the inverse demand and supply equations
ols_demand <- lm(p ~ q + ps + di, data = truffles)
ols_supply <- lm(p ~ q + pf, data = truffles)
summary(ols_demand)
summary(ols_supply)
```

**Explanation.**
Estimating the inverse equations by OLS ignores the simultaneity between $P$ and $Q$, so `q` is correlated with the error term and the OLS estimates are biased and inconsistent. The signs are broadly the same as the 2SLS estimates, but the magnitude of the slope on `q` differs because OLS does not correct for endogeneity. The 2SLS estimates in part (b) are the preferred ones.
