---
title: "Financial Econometrics - Exercise 3.31"
author: "Tony"
date: "2026-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(POE5Rdata)
# library(ggplot2) 先放著
data(tuna)
```

## Exercise 3.31: Demand Analysis for Canned Tuna
This study analyzes 52 weeks of sales data for a major brand of canned tuna.
We explore the relationship between unit sales (SAL1) and the price per can (APR1).

## Part (a): Summary Statistics and Time Series Analysis
**Answer:**

* **Summary Statistics:** The sample mean for weekly sales is 6718.7 units with a very high standard deviation of 6915.1, indicating significant volatility. In contrast, the price per can (APR1) is relatively stable, averaging $0.78 with a standard deviation of only $0.098.

* **Time Series Observation:** Plotting these variables against time (WEEK) reveals that sales spikes often coincide with price drops. This suggests that consumers are responsive to promotional pricing throughout the year.

```{r}
# --- (a) Summary Statistics & Time Series Plots ---
# 計算 SAL1 與 APR1 的平均、最小、最大、標準差

summary_stats <- data.frame(
  Variable = c("SAL1", "APR1"),
  Mean = c(mean(tuna$sal1), mean(tuna$apr1)),
  Min = c(min(tuna$sal1), min(tuna$apr1)),
  Max = c(max(tuna$sal1), max(tuna$apr1)),
  SD = c(sd(tuna$sal1), sd(tuna$apr1))
)
print(summary_stats)

# 繪製隨時間變動的圖(時間序列圖)
###取消 僅學習par(mfrow = c(2, 1)) # 分成上下兩格圖###
plot(1:52, tuna$sal1, type = "l", col = "blue", main = "Weekly Sales (SAL1)"
     , xlab="Week", ylab="Sales")
plot(1:52, tuna$apr1, type = "l", col = "red", main = "Weekly Price (APR1)"
     , xlab="Week", ylab="Price")
###取消 僅學習par(mfrow = c(1, 1)) # 恢復單圖模式###
```


## Part (b): Sales vs. Price Relationship
**Answer:**

* **Visual Relationship:** The scatter plot shows a clear inverse (negative) relationship between price and sales. As the price per can increases, the number of units sold tends to decrease.

* **Economic Expectation:** This result is exactly as expected under the Law of Demand, which states that, ceteris paribus, the quantity demanded of a good falls when the price of the good rises.
```{r}
# --- (b) Scatter Plot 散佈圖 ---
plot(tuna$apr1, tuna$sal1, xlab = "Price (APR1)", ylab = "Sales (SAL1)",
     main = "Sales vs Price")
```


## Part (c): Linear Regression and the Effect of Price (in Cents)
**Answer:**

* **Point Estimate:** The coefficient $\beta_2$ represents the estimated change in unit sales for every one-cent increase in the price. Based on our model, a one-cent increase in price leads to an estimated decrease of -434.4473  units sold.

* **95% Interval Estimate:** In repeated sampling, approximately 95% of the intervals constructed using this methodology will contain the true value of the effect of a one-cent price increase on sales. For this sample, the interval is [[-592.2897], [-276.6049]].
```{r}
# --- (c) Linear Regression (Unit: Cents) ---
tuna$price1 <- 100 * tuna$apr1  # 換算成 cent
mod_tuna <- lm(sal1 ~ price1, data = tuna)
summary(mod_tuna)


# 提取 Beta2 估計值與 95% 信賴區間
b2_tuna <- coef(mod_tuna)["price1"]
ci_b2 <- confint(mod_tuna, "price1", level = 0.95)

cat("Point Estimate (Beta2): ", b2_tuna, "\n")
cat("95% CI for Beta2: [", ci_b2[1], ",", ci_b2[2], "]\n")
```


## Part (d): Expected Sales at 70 Cents
**Answer:**

* **Prediction:** For a week where the price is set at 70 cents ($PRICE1 = 70$), the expected number of cans sold is [10302.9].

* **90% Interval Estimate:** Using the principle of repeated sampling, we expect that 90% of such intervals calculated from different samples would capture the true population mean sales at this price point. Our calculated 90% interval is [[8624.934], [11980.87]].
```{r}
# --- (d) Expected Sales at 70 Cents ---
# 用 predict 函數計算 90% 信賴區間
new_data <- data.frame(price1 = 70)
predict_d <- predict(mod_tuna, new_data, interval = "confidence", level = 0.90)
print(predict_d)
```


## Part (e): Price Elasticity of Demand at the Means
**Answer:**

* **Elasticity Estimate:** The estimated price elasticity of demand, calculated at the sample means, is [-5.059825].

* **Interpretation:** This value implies that a 1% increase in price leads to an approximate [5.059825]% decrease in the quantity of tuna sold.

* **95% Interval Estimate:** In repeated sampling, approximately 95% of intervals constructed this way will contain the true price elasticity at the means. Our calculated 90% interval is [[-6.898149], [-3.221501]].

* **Economic Sense:** Since the absolute value of the elasticity is greater than 1, the demand for this brand of tuna is highly elastic. This makes economic sense because canned tuna is a non-essential commodity with many close substitutes (other brands or protein sources), making consumers very price-sensitive.
```{r}
# --- (e) Elasticity at the Means ---
# Elasticity = Beta2 * (mean_p / mean_q)
mean_p <- mean(tuna$price1)
mean_q <- mean(tuna$sal1)
elas_mean <- b2_tuna * (mean_p / mean_q)

# 彈性的標準誤 (假設平均數為常數)
# SE(Elasticity) = SE(Beta2) * (mean_p / mean_q)
se_b2_tuna <- sqrt(vcov(mod_tuna)["price1", "price1"])
se_elas <- se_b2_tuna * (mean_p / mean_q)

# 95% 彈性信賴區間
df_tuna <- df.residual(mod_tuna)
t_crit_95 <- qt(0.975, df_tuna)
lower_elas <- elas_mean - t_crit_95 * se_elas
upper_elas <- elas_mean + t_crit_95 * se_elas

cat("Elasticity at the Means: ", elas_mean, "\n")
cat("95% CI for Elasticity: [", lower_elas, ",", upper_elas, "]\n")
```


## Part (f): Hypothesis Test for Elasticity
**Answer:**

* **Hypotheses:** We test the Null Hypothesis ($H_0: \epsilon = -3$) against the Alternative Hypothesis ($H_1: \epsilon \neq -3$).

* **Test Statistic and P-value:** The calculated t-statistic is [-2.250572] with a corresponding p-value of [0.02884204 ].

* **Conclusion:** Using a 10% significance level, we compare the p-value to $\alpha = 0.10$. If the p-value is less than 0.10, we reject the null hypothesis and conclude that the price elasticity is significantly different from -3. Otherwise, we fail to reject the null.

```{r}
# --- (f) Hypothesis Test: Elasticity = -3 (10% level) ---
# H0: Elasticity = -3, H1: Elasticity != -3
elas_null <- -3
t_stat_f <- (elas_mean - elas_null) / se_elas
p_val_f <- 2 * (1 - pt(abs(t_stat_f), df_tuna))

cat("T-statistic for Hypothesis Test: ", t_stat_f, "\n")
cat("P-value: ", p_val_f, "\n")
```

